Single-fluorophore orientation determination with multiview polarized illumination: modeling and microscope design

### 2017-08-11 paper/paper/paper.tex
Patrick and Rudolf's second round of edits.
1) Revised justification for using polarized illumination in the introduction.
2) Removed all mentions of frames---referred to intensity measurements instead.
3) New Figure 1d) to clarify the geometry and positions of vectors.
4) Uniform color scale for all figures.
5) Rewording.

\begin{document}
\title{Single-molecule orientation determination with multiview polarized
  illumination: modeling and microscope design}

\author{Talon Chandler,\authormark{1,*} Shalin Mehta,\authormark{2} Hari Shroff,\authormark{3,4} Rudolf Oldenbourg,\authormark{2, 5} and Patrick J. La Rivi\`ere\authormark{1, 4}}
\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, Illinois 60637, USA\\
  \authormark{2}Marine Biological Laboratory, Bell Center for Regenerative Medicine, Woods Hole, Massachusetts, USA\\
  \authormark{3}Section on High Resolution Optical Imaging, National Institute
  of Biomedical Imaging and Bioengineering, National Institutes of Health,
  Bethesda, Maryland 20892, USA\\
  \authormark{4}Whitman Center, Marine Biological Laboratory, Woods Hole,
  Massachusetts 02543, USA\\
  \authormark{5}Brown University, Department of Physics, Providence, Rhode
  Island 02912, USA}
\email{*talonchandler@talonchandler.com} %% email address is required

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract*}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single molecules. First, we relate the
  orientation of single molecules to measurable intensities in multiview
  microscopes and develop an information theoretic metric---the solid angle
  uncertainty---to compare the ability of multiview microscopes to estimate the
  orientation of single molecules. Next, we compare a broad class of microscopes
  using this metric---single- and dual-view microscopes with varying
  illumination polarization, illumination numerical aperture (NA), detection NA,
  obliquity, asymmetry, and exposure. We find that multiview microscopes are
  better at determining the orientation of single molecules than single-view
  microscopes. We also find that choosing a small illumination NA and a large
  detection NA are good design choices, that multiview widefield microscopes can
  benefit from oblique illumination and detection, and that asymmetric NA
  microscopes can benefit from exposure asymmetry.
\end{abstract*}

@@

\section{Introduction}
The orientation of single molecules is a valuable reporter for biological
processes. By imaging biological samples with fluorescent probes that are fixed
to known structures, researchers have studied the dynamics of motor proteins
\cite{peterman2001, forkey2003}, DNA \cite{backer2016}, actin \cite{mehta2016},
septin \cite{demay2011, mcquilken2017}, and membranes \cite{anantharam2010}. The
number of techniques available to biologists for measuring the orientation of
single molecules in live cells is continually expanding. A recent review
\cite{backlund2014} identified three major categories of single molecule
orientation determination techniques:

\textbf{Spatial variation of emission} techniques image single molecules and use
the spatial distribution of intensity in the back focal plane \cite{lieb2004} or
in the image plane \cite{backer2014} to estimate their orientation. The best
techniques in this category use defocusing or phase masks in the back focal
plane to encode orientation and position information in the image
\cite{agrawal2012}. These techniques can estimate the orientation of
molecules in all orientations and only require a single frame, but they
require complex reconstruction algorithms, expensive high numerical aperture
(NA) optics, and sensitive detectors.

@@

\textbf{Polarized illumination or detection} techniques exploit the anisotropic
absorption and emission pattern of single molecules to estimate their
orientation. By measuring the total intensity in the detector plane while
varying polarizer orientations on the illumination or detection path, the
orientation of the molecule can be estimated. These techniques are easy to
implement---changing the illumination or detection polarization is simple with a
polarization splitter \cite{mehta2016} or universal compensator
\cite{oldenbourg1995}---and the reconstruction methods are straightforward
\cite{fourkas2001, mehta2016, backer2016}. The main drawback is that these
techniques require several polarization frames to estimate the orientation of a
single molecule. Despite the requirement for multiple frames, polarized
illumination or detection methods are good choices for live cell imaging because
of their good combination of speed and ease of implementation.

The choice remains---is it best to put polarizers on the illumination path or
the detection path? Placing a polarizer in the illumination path lengthens
acquisition time because the polarizer needs to be changed for each polarization
frame, but polarized illumination provides the largest intensity differences
between molecules in different orientations. Placing polarizers in the detection
path doesn't slow down acquisition (multiple polarization frames can be
collected at once with a polarization splitter), but polarized detection
provides a smaller difference in intensity between molecules in different
orientations than polarized illumination. We are most interested in finding the
orientation of molecules with high precision, so polarized illumination is the
method we consider and extend in this paper.

All single molecule orientation determination methods suffer from some degree of
anisotropic orientation uncertainty, i.e., some molecular orientations cannot be
determined as precisely as others. In some cases the orientation cannot be
determined at all---the forward model can be degenerate for specific dipole
orientations \cite{fourkas2001, lu2008}. We feel that the importance of
isotropic orientation uncertainty has been underappreciated. To our knowledge,
the only authors who have considered anisotropic orientation uncertainty have
only analyzed a small set of molecular orientations instead of all orientations
\cite{agrawal2012}. Our view is that an ideal technique for determining the
orientation of single molecules can reconstruct the orientation with a small and
nearly uniform uncertainty for all molecule orientations.

Recently, there has been increasing interest in multiview microscopy techniques
for biological imaging \cite{wu2013, keller2015, wu2016, wu2017}. Multiview
microscopes offer two major advantages over single-view microscopes. First,
multiview microscopes can achieve nearly isotropic resolution compared to the
poor axial resolution of single-view microscopes. Second, multiview microscopes
can use light-sheet illumination to reduce phototoxicity without requiring
dedicated illumination optics. For example, if two orthogonal objectives are
focused on the same point, then both objectives can alternate roles as the
light-sheet illumination path and as the detection path. Light-sheet
illumination offers a major reduction of phototoxicity because a light sheet
only illuminates in focus regions while single-view microscopes illuminate out
of focus regions. Together, these advantages make multiview microscopes good
candidates for imaging live biological specimens for long periods.

In this work we explore the use of polarized illumination in multiview
microscopes for determining the orientation of single molecules. Existing
multiview microscopes can easily be outfitted with fast-switching polarizers
that do not degrade image quality, so polarized illumination is a natural way to
augment multiview microscopes for measuring the orientation of single
molecules. Furthermore, multiview microscopes can achieve nearly isotropic
resolution while delivering selective illumination. We hypothesize that
multiview microscopes will also provide a small and uniform orientation
uncertainty for single molecules in all orientations. In section \ref{methods}
we develop the required theory for polarized illumination microscopy and develop
information theoretic metrics to compare multiview microscopes. In section
\ref{results} we compare the results for a wide range of multiview microscope
designs. Finally, in section \ref{discussion} we discuss the results and their
impact on polarized multiview microscope design.

\section{Methods}\label{methods}
In this section we develop a method to compare multiview polarized illumination
microscopes for the task of reconstructing molecular orientations. In sections
\ref{excitation}--\ref{forward} we develop the forward model for single-frame
microscopes with a fixed illumination and detection configuration. In section
\ref{designs} we list a variety of ways that single-frame microscopes can be
combined to create experimentally realizable multi-frame microscopes. Finally,
in section \ref{metrics} we develop metrics that we will use to compare
multi-frame microscopes.

Note our distinction between a \emph{frame} and a \emph{view}. A frame refers to
a single intensity measurement with a fixed polarizer setting, illumination
path, and detection path. Acquiring additional frames entails varying the
polarizer setting, illumination path, or detection path. A view refers to a
detection path, and multiview microscopes use multiple detection paths.

@@

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{frames}
\caption{Coordinate systems for a) the absorption dipole moment
  $\hat{\bs{\mu}}_{\text{abs}}$, b) the polarizer transmission axis
  $\hat{\mb{e}}_{\text{exc}}$, and c) the dummy integration vector
  $\hat{\mb{r}}$.}
\label{fig:coordinates}
\end{figure}

We consider a single molecule with an absorption dipole moment
$\hat{\bs{\mu}}_{\text{abs}}$ that we express in spherical coordinates $\Theta$
and $\Phi$ (see Figure \ref{fig:coordinates}a)) as
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}}(\Theta, \Phi) = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}} \label{eq:mu}.
\end{align}
We place the molecule at the focal point of an ideal, aplanatic,
polarization-preserving objective with the optical axis aligned with the
$\mh{z}$ axis. Finally, we place a polarizer behind the objective with a
variable transmission axis $\hat{\mb{e}}_{\text{exc}}$ that we express as
\begin{align}
 \hat{\mb{e}}_{\text{exc}}(\phi_{\text{exc}}) = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
polarizer and the positive $\mh{x}$ axis (see Figure
\ref{fig:coordinates}b)). We illuminate the polarizer with a beam of collimated
light propagating along the optical axis, so the electric field incident on the
objective is parallel with the transmission axis of the polarizer.

We denote the illuminated region of the objective by $\Omega$. If we use
unit vectors $\hat{\mb{r}}$ expressed in spherical coordinates $\theta$ and
$\phi$ (see Figure \ref{fig:coordinates}c)) as
  \begin{align}
   \hat{\mb{r}}(\theta, \phi) = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}}, 
  \end{align}
  then the region $\Omega$ can be expressed as
  \begin{align}
  \Omega = \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\},
  \end{align}
  where $\alpha$ is the angle between the optical axis and the most oblique
  illuminating ray. Equivalently, $\alpha$ can be expressed in terms of NA using
  \begin{align}
    \text{NA} = n\sin\alpha, 
  \end{align}
  where $n$ is the index of refraction of the sample medium. Note that the NA of
  the objective is always greater than or equal to the illumination NA because
  we can under fill the back aperture of the objective.

  The objective focuses the collimated illumination by applying a position-dependent
  rotation to the electric field which we model by multiplying the incident electric
  field $\hat{\mb{e}}$ by a position-dependent rotation matrix
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & -\sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & -\sin\theta\sin\phi \\ \sin\theta\cos\phi& \sin\theta\sin\phi & \cos\theta \end{bmatrix} \label{eq:rot_mat}.
  \end{align}

  To find the absorption efficiency $\eta_{\text{abs}}$ of a single molecule if
  it were illuminated by a single ray, we take the unit incident electric field
  $\hat{\mb{e}}$, take the dot product with the absorption dipole moment, then
  take the modulus squared
    \begin{align}
      \eta_{\text{abs}} &= |\hat{\bs{\mu}}_{\text{abs}}\cdot\mh{e}_{\text{exc}}|^2. \label{eq:singleray}
    \end{align}

    To find the absorption efficiency of a single molecule under incoherent
    focused light, we integrate over all rays and divide by the total incident
    power which gives the vector expression
\begin{align}
  \eta_{\text{abs}} = \frac{\int_{\Omega}d\mh{r}|\hat{\bs{\mu}}_{\text{abs}}\cdot\mb{R}(\mh{r})\mh{e}_{\text{exc}}|^2}{\int_{\Omega}d\mh{r}}\label{eq:abs}. 
\end{align}

@@

At the beginning of this section we assumed that the molecule was located at the
focal point of the condenser. If the illumination objective is aplanatic and we
use K\"ohler illumination, then we can relax this assumption and the above
expressions are valid for molecules anywhere in the focal plane.

We also assumed that we used incoherent illumination, but we can extend these
results to light-sheet illumination that uses a scanned coherent laser. If we
illuminate the back aperture with a weakly focused laser beam and scan the laser
beam slowly---if the scan velocity is much less than the beam waist radius
divided by the coherence time---we can ignore longitudinal excitation and the
molecule will be excited as if a single plane wave was incident. Therefore, we
can find the absorption efficiency of a molecule under weakly focused scanned
laser illumination by taking the limit of equation \ref{eq:scalarabs} as
$\alpha \rightarrow 0$ (or by plugging equation \ref{eq:mu} and
\ref{eq:excfield} into equation \ref{eq:singleray}) giving
\begin{align}
  \eta_{\text{abs}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to widefield
illumination with a low NA. Notice that the absorption efficiency can take its
maximum range of values ($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

Note that $A$ and $B$ in equation \ref{eq:coefficients} are a factor of
$\frac{3}{2}$ larger than the expressions given by Fourkas. We found that
Fourkas' expressions were incorrectly normalized (the limit of
$\eta_{\text{det}}$ as $\alpha\rightarrow \pi/2$ should be 1), and the extra
factor of $\frac{3}{2}$ corrects the error. This correction does not affect
Fourkas' orientation reconstruction but it does mean that Fourkas' algorithm
under predicts the total emitted intensity by a factor of $\frac{3}{2}$.

@@

\subsection{Oblique Illumination and Detection}\label{oblique}
In sections \ref{excitation} and \ref{detection} we assumed that both the
illumination and detection objectives had $\mh{z}$ aligned optical axes, i.e.,
the same objective was used for illumination and detection. To extend the
forward model to oblique optical axes which allows for non-coincident
illumination and detection objectives, we express the dipole orientation in
rotated coordinates using the following expressions
\begin{align}
    \Theta' &= \arccos\left(\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta\right)\label{eq:thetap}\\
  \Phi' &=
          \begin{cases}
            \arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &0 \leq \Phi < \pi  \\
            -\arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &-\pi \leq \Phi < 0\\
          \end{cases}
\end{align}
where $\psi$ is the angle of a right handed rotation about the $\mh{y}$ axis
that maps the $\mh{z}$ axis onto the new optical axis.

\subsection{Single-Frame Microscopes}\label{forward}
The detected intensity is the product of the the sample exposure
$I_{\text{tot}}$, the absorption efficiency, and the detection efficiency
\begin{align}
  I = I_{\text{tot}}\eta_{\text{abs}}\eta_{\text{det}}\label{eq:forward_frame}.
\end{align}
Equation \ref{eq:forward_frame} is the forward model for a \emph{single-frame
  microscope}. Figure \ref{fig:single-frame} shows the efficiencies from three
representative examples of single-frame microscopes.

The intensity measured by a single-frame microscope does not give us enough
information to reconstruct the orientation of a single molecule. Next, we
will consider combining several single-frame microscopes to create
\emph{multi-frame microscopes.}

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single-frame microscopes.\newline \newline
  \textbf{Columns left to right:} 1) schematics of single-frame microscopes
  where the solid line encloses the illumination solid angle, the dashed line
  encloses the detection solid angle, and the arrow indicates the transmission
  axis of the illumination polarizer; 2) the absorption efficiency, see equation
  \ref{eq:scalarabs}; 3) the detection efficiency, see equation
  \ref{eq:scalardet}; (4) the total efficiency, the product of the absorption
  and detection efficiencies.\newline \newline \textbf{Rows top to bottom:} 1)
  coincident illumination ($\text{NA} = 1.1$ with $x$-polarized light) and
  detection (NA = 1.1); 2) non-coincident orthogonal illumination
  ($\text{NA} = 0.8$) and detection ($\text{NA} = 0.8$); 3) non-coincident
  135${}^{\circ}$-separated illumination ($\text{NA} = 0.8$) and detection
  ($\text{NA} = 0.8$). All simulations use $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

\subsection{Multi-Frame Microscopes}\label{designs}
One way to collect multiple frames is to add a universal compensator to the
illumination arm and rapidly select the incident polarization by changing
$\phi_{\text{exc}}$ \cite{oldenbourg1995}. All of the multi-frame microscopes we
will consider in this paper use four frames per view with the illumination
polarization separated by 45${}^{\circ}$.

We also consider several dual-view designs. We evaluate two-view designs that
allow for illumination and detection from both objectives (see \cite{wu2013} for
a dual-view design). In all cases we consider the effect of varying the
illumination and detection numerical aperture. We also consider asymmetric NA
designs (see \cite{wu2017} for an asymmetric NA design) and the effect of
asymmetric sample exposures. Note that we use the word \emph{symmetry} to refer
to dual-view designs with objectives that are identical and the word
\emph{asymmetry} to refer to dual-view designs with objectives that have a
different NA or sample exposure.

>>>>

\begin{document}
\title{Single-fluorophore orientation determination with multiview polarized
  illumination: modeling and microscope design}

\author{Talon Chandler,\authormark{1,*} Shalin Mehta,\authormark{2} Hari Shroff,\authormark{3,4} Rudolf Oldenbourg,\authormark{2, 5} and Patrick J. La Rivi\`ere\authormark{1, 4}}
\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, Illinois 60637, USA\\
  \authormark{2}Marine Biological Laboratory, Bell Center, Woods Hole, Massachusetts, USA\\
  \authormark{3}Section on High Resolution Optical Imaging, National Institute
  of Biomedical Imaging and Bioengineering, National Institutes of Health,
  Bethesda, Maryland 20892, USA\\
  \authormark{4}Whitman Center, Marine Biological Laboratory, Woods Hole,
  Massachusetts 02543, USA\\
  \authormark{5}Brown University, Department of Physics, Providence, Rhode
  Island 02912, USA}
\email{*talonchandler@talonchandler.com} %% email address is required

%%%%%%%%%%%%%%%%%%% abstract and OCIS codes %%%%%%%%%%%%%%%%
\begin{abstract*}
  We investigate the use of polarized illumination in multiview microscopes for
  determining the orientation of single fluorescent molecules. First, we relate
  the orientation of single molecules to measurable intensities in multiview
  microscopes and develop an information-theoretic metric---the solid-angle
  uncertainty---to compare the ability of multiview microscopes to estimate the
  orientation of single molecules. Next, we compare a broad class of microscopes
  using this metric---single- and dual-view microscopes with varying
  illumination polarization, illumination numerical aperture (NA), detection NA,
  obliquity, asymmetry, and exposure. We find that multiview microscopes are
  better at determining the orientation of single molecules than single-view
  microscopes. We also find that choosing a small illumination NA and a large
  detection NA are good design choices, that multiview widefield microscopes can
  benefit from oblique illumination and detection, and that asymmetric NA
  microscopes can benefit from exposure asymmetry.
\end{abstract*}

@@

\section{Introduction}
The orientation of single molecules is a valuable reporter for
biological processes. By imaging biological samples with fluorescent probes that
are fixed to known structures, researchers have studied the dynamics of motor
proteins \cite{peterman2001, forkey2003}, DNA \cite{backer2016}, actin
\cite{mehta2016}, septin \cite{demay2011, mcquilken2017}, and membranes
\cite{anantharam2010}. The number of techniques available to biologists for
measuring the orientation of single molecules in live cells is continually
expanding. A recent review \cite{backlund2014} identified three major categories
of single-molecule orientation determination techniques:

\textbf{Spatial/angular variation of emission} techniques image single molecules
and use the distribution of intensity in the back focal plane \cite{lieb2004} or
in the image plane \cite{backer2014} to estimate their orientation. The best
techniques in this category use defocusing or phase masks in the back focal
plane to encode orientation and position information in the image
\cite{agrawal2012}. These techniques can estimate the orientation of molecules
in all orientations and only require a single intensity measurement, but they
require complex reconstruction algorithms, expensive high numerical aperture
(NA) optics, and sensitive detectors.

@@

\textbf{Polarized illumination or detection} techniques vary the polarization of
light in the illumination or detection paths and exploit the anisotropic
absorption and emission patterns of single molecules to estimate their
orientation. By illuminating and detecting from the entire sample, not just the
focal volume, these techniques can estimate the orientation of many molecules in
parallel without scanning. These techniques are easy to implement---changing the
illumination or detection polarization is simple with a polarization splitter
\cite{mehta2016} or universal compensator/polarizer \cite{shribak2003}---and the
reconstruction methods are straightforward \cite{fourkas2001, mehta2016,
  backer2016}. The main drawback is that these techniques require several
intensity measurements under different polarization orientations to estimate the
orientation of a single molecule. Despite the requirement for multiple
measurements, polarized illumination or detection methods are good choices for
live cell imaging because of their combination of speed and ease of
implementation.

To develop a polarized illumination or detection technique we need to choose a
method for varying the polarization and place it in the illumination or
detection path. The simplest and most inexpensive way to vary the polarization
is with a division-of-time (DoT) polarizer using a rotating polarizer or a
universal compensator/polarizer. Other polarization methods
(divison-of-amplitude, division-of-focal-plane) are more expensive and require
additional alignment or interpolation \cite{tyo2006}. The simplicity of DoT
polarization comes at the expense of temporal resolution---intensity
measurements are made one after the other which increases the acquisition
time. If we place the DoT polarizer in the detection path, we will block
valuable photons from reaching the detector which unnecessarily increases
exposure to the sample \cite{demay2011}. By placing the DoT polarizer in the
illumination arm we minimize exposure to the sample. We focus on DoT polarized
illumination in this paper because of its combination of simplicity and low
sample exposure.

All single molecule orientation determination methods suffer from some degree of
anisotropic orientation uncertainty, i.e., some molecular orientations cannot be
determined as precisely as others. In some cases the orientation cannot be
determined at all---the forward model can be degenerate for specific dipole
orientations \cite{fourkas2001, lu2008}. We feel that the importance of
isotropic orientation uncertainty has been underappreciated. To our knowledge,
the only authors who have considered anisotropic orientation uncertainty have
only analyzed a subset of molecular orientations \cite{agrawal2012}. Our view is
that an ideal technique for determining the orientation of single molecules can
reconstruct the orientation with a small and nearly uniform uncertainty for all
molecule orientations.

Recently, there has been increasing interest in multiview microscopy techniques
for biological imaging \cite{wu2013, keller2015, wu2016, wu2017}. Multiview
microscopes offer two major advantages over single-view microscopes. First,
multiview microscopes can achieve nearly isotropic resolution compared to the
anisotropic resolution of single-view microscopes whose axial resolution is
reduced by at least a factor of two compared to lateral resolution. Second,
multiview microscopes can use light-sheet illumination to reduce phototoxicity
without requiring dedicated illumination optics. For example, if two orthogonal
objectives are focused on the same point, then the objectives can alternate
roles as the light-sheet illumination path and as the detection
path. Light-sheet illumination offers a major reduction of phototoxicity because
a light sheet only illuminates in-focus regions while other widefield and
confocal approaches inevitably illuminate out-of-focus regions. Together, these
advantages make multiview microscopes good candidates for imaging live
biological specimens for long periods.

In this work we explore the use of polarized illumination in multiview
microscopes for determining the orientation of single molecules. Existing
multiview microscopes can easily be outfitted with fast-switching polarizers
that do not degrade image quality, so polarized illumination is a natural way to
augment multiview microscopes for measuring the orientation of single
molecules. Furthermore, multiview microscopes can achieve nearly isotropic
resolution while delivering selective illumination. We find that multiview
microscopes also provide a small and uniform orientation uncertainty for single
molecules in all orientations. In section \ref{methods} we develop the theory
required to model polarized illumination microscopes, and we develop metrics to
compare microscope designs. In section \ref{results} we compare the results for
a wide range of multiview microscope designs. Finally, in section
\ref{discussion} we discuss the results and their impact on polarized multiview
microscope design.

\section{Methods}\label{methods}
In this section we develop a method to compare multiview polarized illumination
microscopes for the task of reconstructing molecular orientations. In sections
\ref{excitation}--\ref{forward} we develop a model for the intensity collected
with a fixed illumination and detection configuration as a function of the
molecule orientation. In section \ref{designs} we list a variety of ways to
combine these intensity measurements into experimentally realizable polarized
illumination microscopes. Finally, in section \ref{metrics} we develop metrics
that we will use to compare these microscopes.

@@

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{frames}
\caption{Coordinate systems for a) the absorption dipole moment
  $\hat{\bs{\mu}}_{\text{abs}}$, b) the polarizer transmission axis
  $\hat{\mb{e}}_{\text{exc}}$, and c) the dummy integration vector
  $\hat{\mb{r}}$. d) The scene consists of a single molecule at the focal point
  of an objective lens with a polarizer in the aperture plane.}
\label{fig:coordinates}
\end{figure}

We consider a single molecule with an absorption dipole moment
$\hat{\bs{\mu}}_{\text{abs}}$ that we express in spherical coordinates $\Theta$
and $\Phi$ (see Figure \ref{fig:coordinates}a)) as
\begin{align}
  \hat{\bs{\mu}}_{\text{abs}}(\Theta, \Phi) = \sin\Theta\cos\Phi\hat{\mb{x}} + \sin\Theta\sin\Phi\hat{\mb{y}} + \cos\Theta\hat{\mb{z}} \label{eq:mu}.
\end{align}
We place the molecule in the focal plane of an ideal, aplanatic,
polarization-preserving objective lens with its optical axis aligned with the
$\mh{z}$ axis. Next, we place a polarizer in the aperture plane with a variable
transmission axis $\hat{\mb{p}}_{\text{exc}}$ that we express as
\begin{align}
 \hat{\mb{p}}_{\text{exc}}(\phi_{\text{exc}}) = \cos\phi_{\text{exc}}\hat{\mb{x}} +
\sin\phi_{\text{exc}}\hat{\mb{y}},\label{eq:excfield}
\end{align}
where $\phi_{\text{exc}}$ is the angle between the transmission axis of the
polarizer and the positive $\mh{x}$ axis (see Figure
\ref{fig:coordinates}b)). Finally, we place a spatially incoherent light source
(or its image) in the aperture plane to establish widefield illumination. In
this geometry each point in the sample is illuminated by a range of angles that
can be changed by adjusting the radius of the stop in the aperture plane. We
denote the set of illuminated angles by $\Omega$ (see Figure
\ref{fig:coordinates}d)). If we use unit vectors $\hat{\mb{r}}$ expressed in
spherical coordinates $\theta$ and $\phi$ (see Figure \ref{fig:coordinates}c))
as
  \begin{align}
   \hat{\mb{r}}(\theta, \phi) = \sin\theta\cos\phi\hat{\mb{x}} + \sin\theta\sin\phi\hat{\mb{y}} + \cos\theta\hat{\mb{z}}, 
  \end{align}
  then the set of angles $\Omega$ can be expressed as
  \begin{align}
  \Omega = \{\phi, \theta\ |\ 0 < \phi \leq 2\pi, 0 < \theta \leq \alpha\},
  \end{align}
  where $\alpha$ is the angle between the optical axis and the most oblique
  illuminating ray. Equivalently, $\alpha$ can be expressed in terms of NA using
  \begin{align}
    \text{NA} = n\sin\alpha, 
  \end{align}
  where $n$ is the index of refraction of the sample medium. Note that the NA of
  the objective is always greater than or equal to the illumination NA because
  we can underfill the aperture plane.

  The objective focuses the light by applying a position-dependent rotation to
  the electric field which we model by multiplying the incident electric field
  $\hat{\mb{p}}_{\text{exc}}$ by a position-dependent rotation matrix
  \begin{align}
  \mb{R}(\mh{r}) &= \begin{bmatrix} \cos\theta\cos^2\phi + \sin^2\phi & (\cos\theta -1)\sin\phi\cos\phi & -\sin\theta\cos\phi\\ (\cos\theta - 1)\sin\phi\cos\phi & \cos\theta\sin^2\phi + \cos^2\phi & -\sin\theta\sin\phi \\ \sin\theta\cos\phi& \sin\theta\sin\phi & \cos\theta \end{bmatrix} \label{eq:rot_mat}.
  \end{align}

  To find the absorption efficiency $\eta_{\text{abs}}$ of a single molecule if
  it were illuminated by a single ray, we take the unit incident electric field
  $\hat{\mb{p}}_{\text{exc}}$, take the dot product with the absorption dipole
  moment, then take the modulus squared
    \begin{align}
      \eta_{\text{abs}}^{\text{ray}} &= |\hat{\bs{\mu}}_{\text{abs}}\cdot\mh{p}_{\text{exc}}|^2. \label{eq:singleray}
    \end{align}

    To find the absorption efficiency of a single molecule under widefield
    illumination, we integrate over all illumination rays and divide by the
    total incident power which gives the vector expression
\begin{align}
  \eta_{\text{abs}} = \frac{\int_{\Omega}d\mh{r}|\hat{\bs{\mu}}_{\text{abs}}\cdot\mb{R}(\mh{r})\mh{p}_{\text{exc}}|^2}{\int_{\Omega}d\mh{r}}\label{eq:abs}. 
\end{align}

@@

We can extend these results to light-sheet illumination created by scanning a
laser. If we shine a narrow collimated laser beam onto a galvanometer in the
aperture plane and scan the galvanometer, we create a weakly focused light sheet
in the sample. If we underfill the aperture plane and scan the beam in the
sample with a velocity much less than the beam waist radius divided by the
coherence time, we can ignore longitudinal excitation and molecules in the
sample will be excited as if a single plane wave was incident. Therefore, we can
find the absorption efficiency of a molecule under weakly focused scanned laser
illumination by taking the limit of equation \ref{eq:scalarabs} as
$\alpha \rightarrow 0$ (or by plugging equation \ref{eq:mu} and
\ref{eq:excfield} into equation \ref{eq:singleray}) giving
\begin{align}
  \eta_{\text{abs}}^{\text{ray}} &= \sin^2\Theta\cos^2(\Phi - \phi_{\text{exc}}),
\end{align}
which is recognizable as Malus' law generalized to three dimensions. This means
that light-sheet illumination is approximately equivalent to low-NA widefield
illumination with respect to the molecular orientations they excite. Notice that
the absorption efficiency can take its maximum range of values
($\eta_{\text{abs}} \in [0, 1]$) when $\alpha=0$.

@@

Note that $A$ and $B$ in equation \ref{eq:coefficients} are a factor of
$\frac{3}{2}$ larger than the expressions given by Fourkas. We found that
Fourkas' expressions were incorrectly normalized (the limit of
$\eta_{\text{det}}$ as $\alpha\rightarrow \pi/2$ should be 1), and the extra
factor of $\frac{3}{2}$ corrects the error. This correction does not affect
Fourkas' orientation reconstruction but it does mean that Fourkas' algorithm
underpredicts the total emitted intensity by a factor of $\frac{3}{2}$.

@@

\subsection{Oblique Illumination and Detection}\label{oblique}
In sections \ref{excitation} and \ref{detection} we assumed that both the
illumination and detection objectives had $\mh{z}$ aligned optical axes, i.e.,
the same objective was used for illumination and detection. To extend the
forward model to oblique optical axes which allows for non-coincident
illumination and detection objectives, we express the dipole orientation in a
rotated coordinate frame using the following expressions
\begin{align}
    \Theta' &= \arccos\left(\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta\right)\label{eq:thetap}\\
  \Phi' &=
          \begin{cases}
            \arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &0 \leq \Phi < \pi  \\
            -\arccos\left(\dfrac{\cos\psi\cos\Phi\sin\Theta - \sin\psi\cos\Theta}{\sqrt{1 - (\sin\psi\cos\Phi\sin\Theta + \cos\psi\cos\Theta)^2}}\right) \ \ \ &-\pi \leq \Phi < 0\\
          \end{cases}
\end{align}
where $\psi$ is the angle of a right handed rotation about the $\mh{y}$ axis
that maps the $\mh{z}$ axis onto the new optical axis.

\subsection{Intensity Measurements}\label{forward}
The detected intensity is the product of the the sample exposure
$I_{\text{tot}}$, the absorption efficiency, and the detection efficiency
\begin{align}
  I = I_{\text{tot}}\eta_{\text{abs}}\eta_{\text{det}}\label{eq:forward_frame}.
\end{align}
Equation \ref{eq:forward_frame} is the forward model for a single intensity
measurement. Figure \ref{fig:single-frame} shows the efficiencies in three
representative geometries.

A single intensity measurement does not give us enough information to
reconstruct the orientation of a single molecule. Next, we will consider
combining several intensity measurements to create complete polarized
illumination microscopes.

\begin{figure}[H]
\centering\includegraphics[width=\textwidth]{single-frame}
\caption{Representative examples of single intensity measurements.\newline
  \newline \textbf{Columns left to right:} 1) schematics where the solid line
  encloses the illumination solid angle, the dashed line encloses the detection
  solid angle, and the arrow indicates the transmission axis of the illumination
  polarizer; 2) the absorption efficiency, see equation \ref{eq:scalarabs}; 3)
  the detection efficiency, see equation \ref{eq:scalardet}; (4) the total
  efficiency, the product of the absorption and detection efficiencies.\newline
  \newline \textbf{Rows top to bottom:} 1) coincident illumination
  ($\text{NA} = 1.1$ with $x$-polarized light) and detection (NA = 1.1); 2)
  non-coincident orthogonal illumination ($\text{NA} = 0.8$) and detection
  ($\text{NA} = 0.8$); 3) non-coincident 135${}^{\circ}$-separated illumination
  ($\text{NA} = 0.8$) and detection ($\text{NA} = 0.8$). All simulations use
  $n=1.33$.}
  \label{fig:single-frame}
\end{figure}

\subsection{Polarized Illumination Microscope Configurations}\label{designs}
One way to collect multiple intensity measurements is to add a universal
compensator/polarizer to the illumination arm and rapidly select the incident
polarization by changing $\phi_{\text{exc}}$ \cite{shribak2003}. All of the
microscopes we will consider in this paper use four polarizations per
illumination path with the polarization orientations separated by
45${}^{\circ}$.

We also consider several dual-view designs. We evaluate dual-view designs that
allow for illumination and detection from both objectives (see \cite{wu2013} for
a dual-view design). In all cases we consider the effect of varying the
illumination and detection numerical aperture. We also consider asymmetric NA
designs (see \cite{wu2017} for an asymmetric NA design) and the effect of
asymmetric sample exposures. Note that we use the word \emph{symmetry} to refer
to dual-view designs with identical objectives and sample exposures and the word
\emph{asymmetry} to refer to dual-view designs with objectives that have a
different NA or sample exposure.
